What we {\sf eval} in the shadows

### 2021-04-01 paper/main.tex
format

\mypara{Base.}
???

@@

This section focuses on our corpus of \Corpus CRAN packages that use
\eval; these represent {\bf 28.9\%} of the entire repository. To provide % \Corpus / \CranAll
a picture of \eval usage, all runnable code in CRAN was executed. Runs
that did not exercise an \eval in the corpus were discarded. We
retained \Nbruns to generate the data presented here. In total,
\Allcalls invocations of \eval and its variants were recorded. Out of
the \Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
packages were observed calling it in our runs --- the difference can
be chalked down to incomplete code coverage and occasional
infrastructure failures.

Figure~\ref{freq} summarizes the frequency of calls to \eval over all
runs. The left column is the number of observed calls, the right
column is the number of packages falling in that range. There are
\Fewcalls packages with low \eval frequency, i.e. fewer than 100
calls; and a similar number of packages, \Manycalls to be precise,
that use \eval more than 1,000 times. One package, \Maxcallspack,
calls it \Maxcalls times and thus accounts for over half of the
observed data.

@@

\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 0 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
\tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
\tt 101 &--& \tt 1K    & \Binc  & \tt 1M &--&\tt 10M   & \Bing\\
\tt 1K &--& \tt 10K    & \Bind  & \tt 10M &--& \tt 100M & \Binh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}
\begin{tabular}{|@{\,}r@{\,}|@{\,}rrrr@{\,}|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
\small Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
\small Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}
\end{tabular}
\end{figure}

@@

\medskip  (a) All

@@

\vspace{-2mm}
    \caption{Normalized invocations} \label{cn}
\vspace{-2mm}
\end{wrapfigure}

Figure~\ref{tab:variantseval} illustrates the use of variants of
\eval. For each of the four variants, the first rows shows the number
of call sites occurring in the corpus (\emph{static}), the second row
is the number of call sites that were encountered during tracing
(\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The vast majority of call sites, and calls, are
to \eval itself, \c{eval.parent} is rare, as for \c{evalq} and
\c{local} are barely used at all. The difference between static and
exercised reflects spotty code coverage and is a limitation
of dynamic analysis.

Figure~\ref{cn}(a) shows normalized invocation counts. The column on
the left is the average number of calls from a particular site for any
given run (\emph{\#calls}). On the right is the count of sites that
fall in that range (\emph{\#sites}). For instance, \Runbinh sites are
invoked on average over 2,001 times per run. Larger numbers suggest
uses from within a loop or a recursive context -- this seems to be the
exception as most \evals, \Runbina, are exercised 50 times or less.
Figure~\ref{cn}(b) zooms in on low frequency call sites. The x-axis
shows normalized invocations and the y-axis is the number sites for
that value. Most low-frequency \evals are invoked only once, about
half as many are invoked twice; after that the frequency quickly
drops.

Arguments to \eval can be either code or values, in the latter case
\eval has no effect. Code accounts for \Codepercent of arguments in
our corpus. More specifically, \Symbolpercent of arguments are
\c{symbol}s (i.e. these are single variables such as \c{x}),
\Languagepercent are \c{language} object (an internal R type roughly
corresponding to a single expression, such as \c{x+1}), and
\Expressionpercent are \c{expression} objects (multiple statements).
Further inspection reveals that most symbols, \Ggplotsymbolpercent to
be exact, come from a single call in the \c{ggplot2} package and have
have the value \c{\_inherit}. Even if we remove this outlier, the
majority of \evals are simple expressions.

To estimate how much executable code is dynamically injected into the
program through \eval, we measure the size of arguments in number of
abstract syntax tree nodes as returned by R's parser; for example,
\c{x+1} counts as 3 nodes.\footnote{We experimented with measuring
code size by de-parsing the expressions to strings, but these
measurements were dominated by de-parsed data objects, these could
range in the MBs.} The median argument size is \Medianszeval (recall
the large number of symbols) and the average is \Avgszeval nodes. The
largest \eval input observed is \Maxszeval\,-- a significant chunk of
code. Figure~\ref{fig:sizedistribution} shows the distribution of
sizes for arguments of fewer or equal to 25 nodes. The x-axis is the
size of arguments in number of nodes, and the y-axis is the count of
arguments with that size. The size drops rapidly, with few
observations larger than 15 nodes. The long tail is omitted from the
graph for legibility.

@@

To measure the work performed in \evals, we count the instructions
executed by interpreter during evaluation of the argument.  Most invocations perform relatively little
work with \Smalleventspct of \evals executing 50 or fewer
instructions. The violin plot of Figure~\ref{ev}(a) corresponds to
\evals executing $\leq$ 50 instructions, it is dominated by trivial
symbol look ups. Figure~\ref{ev}(b) has the work-intensive \evals
which go all the way to \MaxeventsRnd instructions.

\begin{figure}[h!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Instructions per packages} \label{ev}
\end{figure}

\paragraph{Base} The base dataset contains 13 packages. All of them trigger at least one \eval call. In total, \baseAllcalls \eval  were recorded. Most of them are calls to the main \eval. There are a few calls to the other variants, less than 4\%. Contrary to the CRAN dataset, there are calls to \c{evalq} (\baseEvalqs). Out of \baseStaticeval \eval static sites, \baseTriggeredeval are exercised, i.e. \baseTriggeredevalpct, out of \baseStaticevalq, \baseTriggeredevalq are exercised, and out of \baseStaticevalparent, \baseTriggeredevalparent are exercised.  Most of the arguments, \baseCodepercent,  are code objects. However, contrary to CRAN, few arguments, only \baseSymbolpercent, are symbols, but are rather language objects, \baseLanguagepercent. There are also few expression objects: \baseExpressionpercent. The prevalence of language objects suggests that \emph{base} mainly performs metaprogramming with \eval.
The median argument size is \baseMedianszeval, which is more than for CRAN, which was dominated by symbol evaluation. The maximum size is \baseMaxszeval. Small event evals (less than 50 events) amount for \baseSmalleventspct of the events inside \eval calls.

% Figures for evalq are weird: more triggered than static call sites

\paragraph{Kaggle} In the Kaggle dataset, \kaggleNbruns notebooks trigger at least one \eval. In total, \kaggleAllcalls \eval were recorded. All of them are calls to the normal \eval.
Out of \kaggleStaticeval static \eval sites, we exercise \kaggleTriggeredeval of them. Most \eval are invoked only once. Only one \eval amounts for several thousand, \kaggleMaxcalls,  invocations. Code amounts for \kaggleCodepercent of arguments in the kaggle corpus. However, contrary to the CRAN corpus, they are all expressions, and there are very few symbols (\kaggleSymbolpercent) and no language objects. Most expression objects result from a call to the \c{parse} function, suggesting that most \eval in Kaggle parse and evaluate strings. The median argument size is \kaggleMedianszeval, which makes sense, as there are very few arguments that are only symbols. The largest argument size is \kaggleMaxszeval. The distribution of events is similar to the one for CRAN or \emph{base}: \kaggleSmalleventspct of \eval calls amount for less than 50 events.

\paragraph{Discussion} \emph{Base} is more akin to using \eval for metaprogramming, \emph{kaggle} to evaluating parsed strings, and CRAN is dominated by symbol evaluation. In terms of number of calls, a few calls dominate, with a call from \emph{ggplot} being prominent in the CRAN corpus. When looking at the \eval sites, \eval is more spread out. Those two measures emphasize two different points of view:

\begin{itemize}
	\item The number of calls is relevant to an optimizing compiler: the most frequent calls are the ones to be optimized first.
	\item The number of sites is relevant to the programmer/user: it describes better how programmers use \eval.
\end{itemize}

% What about a threat model?

@@

The previous section gave a quantitative view of the usage of \eval;
in this section we try to elucidate \emph{what} it does.

@@

\begin{table}[h]
	\begin{tabular}{|c|c|c|c|c|c|}
		\hline
		Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\
		\hline
		\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvirnonparenta \\
		\hline
		\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb \\
		\hline
		\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirnonparentc \\
		\hline
		\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\
		\hline
		\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\
		\hline
		\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\
		\hline
		\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\
		\hline
		\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\
		\hline
		\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\
		\hline
		\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\
		\hline
		\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\
		\hline
	\end{tabular}
	\caption{Minimized expressions: eleven most frequent in number of sites. \#sites is the number of sites; \%sites the percentage of sites across the total number of sites (\packageNbCallSites sites) with at least this minimized expression; \#packages is the number of packages with at least one of the minimized expression and measures how widespread a minimized expression is. \#operations is the average number of steps performed by the R interpreter when given that minimized expression, and \%envir is the percentage of call sites for which the \c{envir} argument is explicitly passed, and is not \c{parent.frame()}, which is the most common explicit environment passed to \eval.} \label{tab:minimizedexpressions}
\end{table}

>>>>

\mypara{Base.} The base dataset contains 13 libraries performing basic
arithmetics, statistics, and operating system related functionalities.

@@

This section focuses on CRAN packages and reports statistics about the usage of
\eval. Calls to \eval appear in {\bf 28\%} of the packages hosted on CRAN. There
are \Corpus packages with an \eval, all programs extracted from these packages
were executed, any run that did not exercise one of the \eval call sites in the
packages was discarded. This left \Nbruns runs of some of the \Corpus packages
to generate the data presented here. There were \Allcalls invocations of \eval
and its variants occurring in \Triggeredpkgs packages.\footnote{Not all packages
with static call sites of \eval were observed triggering that call site, this
can be chalked down to incomplete code coverage and occasional infrastructure
failures of our analysis infrastructure.} Figure~\ref{freq} summarizes the
frequency of dynamic calls to \eval with the left column being the number of
calls and the right, the number of packages in that range. There are \Fewcalls
packages with low \eval frequency, fewer than 100 calls; and a similar number of
packages, \Manycalls to be precise, that use \eval more than 1,000 times. One
package, \Maxcallspack, calls it \Maxcalls times and thus accounts for over half
of the observed calls.

@@

\begin{tabular}{@{}l@{\hspace{1.5cm}}l@{}}
\begin{minipage} {5cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 1 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
\tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
\tt 101 &--& \tt 1K    & \Binc  & \tt 1M &--&\tt 10M   & \Bing\\
\tt 1K &--& \tt 10K    & \Bind  & \tt 10M &--& \tt 100M & \Binh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{minipage}
&
\begin{minipage}{7cm}
\begin{tabular}{|@{\,}r|rrrr|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
\small Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
\small Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
\end{tabular}~\\[2mm]\caption{Variants}\label{tab:variantseval}
\end{minipage}\end{tabular}\end{figure}

@@

  \medskip  (a) All
  \medskip
  \medskip

@@

\vspace{-2mm} \caption{Normalized invocations} \label{cn}\vspace{-2mm}
\end{wrapfigure}

Figure~\ref{tab:variantseval} enumerate the use of variants of \eval. For each
of the four variants, the first rows shows the number of call sites in the
corpus (\emph{static}), the second row is the number of call sites that were
encountered during analysis (\emph{exercised}), the last row is the number of
calls (\emph{invocations}). The overwhelming majority of call sites and dynamic
calls, are to \eval itself, \c{eval.parent} is rare, and both \c{evalq} and
\c{local} are barely used at all. The difference between sites and exercised
sites underscores the limitations due to code coverage.

Figure~\ref{cn}(a) shows normalized invocation counts per call site; on the left
are average number of calls from a given site and given run, on the right are
counts of sites that fall in that range. For instance, \Runbinh sites are
invoked on average over 2,000+ times per run. Larger numbers suggest loops or
recursive contexts -- this seems to be the exception as most \evals, \Runbina,
are exercised 50 times or less. Figure~\ref{cn}(b) zooms in on low frequency
call sites. The x-axis shows normalized invocations and the y-axis is the number
sites for that value. Most low-frequency \evals are invoked only once, about
half as many are invoked twice; after that the frequency quickly drops.

The \eval can be passed any value, but if the argument is not an expression,
\eval returns it without doing anything. Expressions account for \Codepercent of
arguments in our corpus. More specifically, \Symbolpercent of arguments are
\c{symbol}s (single variables such as \c{x}), \Languagepercent are \c{language}
object (a single expression, such as \c{x+1}), and \Expressionpercent are
\c{expression} objects (multiple statements). Further inspection reveals that
most symbols, \Ggplotsymbolpercent to be exact, come from a single call in the
\c{ggplot2} package and have have the value \c{\_inherit}. Even if we remove
this outlier, the majority of \evals are simple expressions.

To estimate how much executable code is injected through \eval, we measure the
the number of abstract syntax tree nodes as returned by the parser; for example,
\c{x+1} counts as 3.\footnote{We attempted to measure string lengths after
de-parsing expressions, but these measurements were dominated by de-parsed data
objects, which could range in the MBs.} The median argument size is
\Medianszeval (due to the many symbols) and the average is \Avgszeval nodes. The
largest \eval input observed is \Maxszeval\,-- a significant chunk of code.
Figure~\ref{fig:sizedistribution} shows the distribution of sizes for arguments
of fewer or equal to 25 nodes. The x-axis is the size of arguments in number of
nodes, and the y-axis is the count of arguments with that size. The size drops
rapidly, with few observations larger than 15 nodes. The long tail is omitted
from the graph for legibility.

@@

To estimate the work performed in \evals, we count the instructions executed by
interpreter. Most invocations perform relatively little work with
\Smalleventspct of \evals executing 50 or fewer instructions. The violin plot of
Figure~\ref{ev}(a) corresponds to \evals executing $\leq$ 50 instructions, it is
dominated by trivial symbol look ups. Figure~\ref{ev}(b) has the work-intensive
\evals which go all the way to \MaxeventsRnd instructions.

\begin{figure}[h!]
\begin{tabular}{@{}c@{}c@{}}
\begin{minipage}{7.5cm}
 \includegraphics[width=\textwidth]{events_per_pack_small}
\end{minipage}&\begin{minipage}{7.5cm}
  \includegraphics[width=\textwidth]{events_per_pack_large}
\end{minipage}\\[-3mm]
\small (a) Small & \small (b) Large
\end{tabular}
 \caption{Instructions per \eval} \label{ev}
\end{figure}

\paragraph{Base}
We recorded \baseAllcalls \eval calls in the base libraries with fewer than 4\%
of those being to any variants. Unlike CRAN, Base calls \c{evalq} (\baseEvalqs).
Out of \baseStaticeval call sites, \baseTriggeredevalpct are exercised. Most
arguments, \baseCodepercent, are expressions. However, unlike CRAN,
\baseLanguagepercent of the arguments are language objects. The median argument
size is \baseMedianszeval, which is more than for CRAN. The maximum size is
\baseMaxszeval. Small instruction counts, less than 50, amount for
\baseSmalleventspct of calls.

\paragraph{Kaggle}
Only \kaggleNbruns scripts trigger \eval. In total, \kaggleAllcalls \eval were
recorded; none are calls to variants. Out of \kaggleStaticeval call sites,
\kaggleTriggeredeval were exercised. Most call sites are invoked only once. Only
one site is called \kaggleMaxcalls times. Expressions amount for
\kaggleCodepercent of arguments. Unlike CRAN corpus, there are very few symbols
(\kaggleSymbolpercent). Most expressions result from calls to \c{parse}, thus
most \evals start with strings. The median argument size is \kaggleMedianszeval,
Which makes sense, as few arguments are only symbols. The largest argument is
\kaggleMaxszeval. The distribution of instructions per eval is similar to CRAN;
\kaggleSmalleventspct of \evals execute fewer than 50 instructions.

\paragraph{Discussion}
These results show that \eval is central to the language implementation as it is
omnipresent in the relatively small Base library. Packages, which are developed
by experienced programmers, make regular and varied use of the facility.
Finally, scripts are written by less sophisticated users, and likely perform
simpler tasks, and thus have lesser need for \eval, and its uses originate from
strings which are easier to manipulate for end-users.

@@

The previous section gave a quantitative view of \eval usage; we now try to
elucidate \emph{what} it does.

@@

\begin{table}[h]\small
\begin{tabular}{|c|c|c|c|c|c|}\hline
Minimized & \#sites & \%sites & \#packages & \#operations & \%envir \\	\hline
\texttt{X} & \packageMinimizedcallsitesa  & \packageMinimizedpropsitesa & \packageMinimizedpackagea & \packageMinimizedoperationsaRnd & \packageMinimizedpercentenvirnonparenta \\\hline
\texttt{F(F())} & \packageMinimizedcallsitesb  & \packageMinimizedpropsitesb & \packageMinimizedpackageb  & \packageMinimizedoperationsbRnd & \packageMinimizedpercentenvirnonparentb \\\hline
\texttt{V} & \packageMinimizedcallsitesc & \packageMinimizedpropsitesc & \packageMinimizedpackagec & \packageMinimizedoperationscRnd & \packageMinimizedpercentenvirnonparentc \\\hline
\texttt{F(X)}& \packageMinimizedcallsitesd & \packageMinimizedpropsitesd & \packageMinimizedpackaged & \packageMinimizedoperationsdRnd & \packageMinimizedpercentenvirnonparentd \\\hline
\texttt{model.frame} & \packageMinimizedcallsitese & \packageMinimizedpropsitese & \packageMinimizedpackagee & \packageMinimizedoperationseRnd & \packageMinimizedpercentenvirnonparente \\\hline
\texttt{F()}& \packageMinimizedcallsitesf & \packageMinimizedpropsitesf & \packageMinimizedpackagef & \packageMinimizedoperationsfRnd & \packageMinimizedpercentenvirnonparentf \\\hline
\texttt{\$}& \packageMinimizedcallsitesg & \packageMinimizedpropsitesg & \packageMinimizedpackageg & \packageMinimizedoperationsgRnd & \packageMinimizedpercentenvirnonparentg \\\hline
\texttt{FUN} & \packageMinimizedcallsitesh & \packageMinimizedpropsitesh & \packageMinimizedpackageh & \packageMinimizedoperationshRnd & \packageMinimizedpercentenvirnonparenth \\\hline
\texttt{<-} & \packageMinimizedcallsitesi  & \packageMinimizedpropsitesi & \packageMinimizedpackagei & \packageMinimizedoperationsiRnd & \packageMinimizedpercentenvirnonparenti \\\hline
\texttt{\{BLOCK\}} & \packageMinimizedcallsitesj & \packageMinimizedpropsitesj & \packageMinimizedpackagej & \packageMinimizedoperationsjRnd & \packageMinimizedpercentenvirnonparentj \\\hline
\texttt{F(F(<-))} & \packageMinimizedcallsitesk & \packageMinimizedpropsitesk & \packageMinimizedpackagek & \packageMinimizedoperationskRnd & \packageMinimizedpercentenvirnonparentk \\\hline
\end{tabular}
\caption{Minimized expressions: eleven most frequent in number of sites. \#sites
  is the number of sites; \%sites the percentage of sites across the total
  number of sites (\packageNbCallSites sites) with at least this minimized
  expression; \#packages is the number of packages with at least one of the
  minimized expression and measures how widespread a minimized expression is.
  \#operations is the average number of steps performed by the R interpreter
  when given that minimized expression, and \%envir is the percentage of call
  sites for which the \c{envir} argument is explicitly passed, and is not
  \c{parent.frame()}, which is the most common explicit environment passed to
  \eval.} \label{tab:minimizedexpressions}
\end{table}
